# Getting Started {#DataManip3}

```{r tidyr3, echo = FALSE, message = FALSE, warning = FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy = FALSE)

```

## Data Download {#DataManip3.1}

The NOS data are downloaded using the `naturecounts` R package. Detailed instruction on how to use the `naturecounts` R pacakge can be found in [NatureCounts: An Introductory Tutorial](link to be provided) or on the [github](https://birdstudiescanada.github.io/naturecounts/index.html) repository. These instructions are not repeated here. 

To access the NOS collections you must [sign up](https://www.birdscanada.org/birdmon/default/register.jsp) for a **free** account and request permission from the data custodian to access each `collection`. Once assess has been granted to can retrieve the data using the `request_id` provided. Also, don't forget to enter your personal `username` below. You will be promted for your password. 

There are many build in options which enable you to select temporal and/or spatial filters prior to download. If you want targeted data, you should add filters to the code chunk below. Because we will need additional auxially data from the [Bird Monitoring Data Exchange] (https://www.birdscanada.org/birdmon/default/nc_bmde.jsp) (BMDE) table in NatureCounts, we will want to change the `field_set` to "extend and the `username` to your own account. 

```{r downloaddata, echo = TRUE, message = FALSE}

#Example code to download individual collections
#raw.data<-nc_data_dl(collections = "ATOWLS", fields_set="extended", username ="dethier", info ="data download NOS")

#Example code to download ALL collections (still need to add ABOWLS when it become available)
raw.data<-nc_data_dl(collections = c("ATOWLS", "BCOWLS", "MBOWLS", "ONOWLS", "QCOWLS", "SKOWLS"), fields_set="extended", username ="dethier", info ="data download NOS")

```

You should save a local copy of the dataset so that you don't need to pull it from the server again. 

```{r savedata, echo = FALSE, message = FALSE}

#write table to working directory so that you have a local copy saved
write.csv(raw.data, "raw.owls.data.csv")
            
#to read saved datafile into R ensure that you 
raw.data<-read.csv("raw.owls.data.csv")

```

## Select {#DataManip3.2}

Now that the data are downloaded we will want to select the columns needed for the analysis. You may not have all the variables below if you didn't change the `field_set` to "extend". That is OK! You may not need all the auxially data for your analytical purposes. 

```{r dataselect1, echo = TRUE, message = FALSE}

in.data<-raw.data %>% select(SamplingEventIdentifier, SurveyAreaIdentifier,RouteIdentifier, SiteCode, collection, survey_day, survey_month, survey_year, survey_week, ProtocolCode, CollectorNumber, EffortUnits1, EffortMeasurement1, EffortUnits3, EffortMeasurement3, EffortUnits5, EffortMeasurement5, EffortUnits11, EffortMeasurement11, EffortUnits14, EffortMeasurement14, species_id, CommonName, ScientificName, latitude, longitude, bcr, StateProvince, ObservationDescriptor, ObservationCount, ObservationDescriptor2, ObservationCount2,ObservationDescriptor3, ObservationCount3)

```

# Filter {#DataManip3.3}

Filter out the data we know we don't want before proceeding. 
 
```{r datafilter1, echo = TRUE, message = FALSE}

#Drop Newfoundland & Labrador based on StateProvince 
in.data<-in.data %>% filter (StateProvince!="Newfoundland") %>% droplevels()

#Drop Northwest Terriroties based on route identifier
in.data<-filter(in.data, !grepl("NT", RouteIdentifier)) %>%  droplevels()

#Remove surveys with missing Month, Day, Year
in.data<-in.data %>% filter (!is.na(survey_day), !is.na(survey_month), !is.na(survey_year))

#Some Ontario Routes only have lat/long for the start point.
#Remove surveys with missing lat long
#in.data<-in.data %>% filter (!is.na(latitude), !is.na(longitude))

#Remove survyes with missing protocol ID 
in.data<-in.data %>% filter (!is.na(ProtocolCode))

#Remove surveys with NOCTOWLS protocol ID
in.data<-in.data %>% filter (ProtocolCode != "NOCTOWLS")

```

If you are using the BC data and want to assign `ProtocolCode` to differentiate regions.  

```{r protocol filter, echo = TRUE, message = FALSE}
#First need to assing BC new protocol ID since this is not done in the underlying database to reflect difference in data collection between region.

#Use this previously loaded table: Regions_BCY.csv

in.data<-left_join(in.data, BCregion, by="RouteIdentifier")
in.data$ProtocolCode <- ifelse(in.data$collection=="BCOWLS", in.data$protocol_id_new, in.data$ProtocolCode)
in.data<-in.data %>% select(-Timing.Region, -protocol_id_new)


```

You may want to fix some data inconsistencies in StateProvince naming. However, you may want to use `collection` or `ProtocolCode` rather than `StateProvince` to do the analysis since there are some points that cross the provincal boundaries. Something to consider. 

```{r datafilter2, echo = TRUE, message = FALSE}

in.data$StateProvince[in.data$StateProvince  == "Ontario"]  <-  "ON"
in.data$StateProvince[in.data$StateProvince  == "British Columbia and Yukon"]  <-  "BCY"
in.data$StateProvince[in.data$StateProvince  == "ME"]  <- "QC"
in.data$StateProvince[in.data$StateProvince  == "NL"]  <- "QC"

#Some of the MBOWL dataset is assigned to SK. May be correct? Left this for now. Will check during plotting. 
```

Filter out the speices that we don't want (e.g., Woodcock, Snipe), but keep the NA because these are representative of a zero-count at a given stop. This step may not be necessary.  

```{r filterspecies, echo = TRUE, message = FALSE}

##THIS IS CAUSING AN ERROR##
#in.data <- in.data %>% filter (CommonName %in% c("Barred Owl", "Boreal Owl", "Great Gray Owl", "Great Horned Owl", "Northern Saw-whet Owl", "Eastern Screeh-Owl", "Western Screech-Owl", "Long-eared Owl", "Short-eared Owl", "Northern Hawk Owl", "Northern Pygmy-Owl", "Snowy Owl", "Barn Owl", "Flammulated Owl") | is.na(CommonName))

```

Next we add a day-of-year column using the `format_dates` [helper function](https://birdstudiescanada.github.io/naturecounts/reference/format_dates.html).  

```{r format dates, echo = TRUE, message = FALSE}

#create a doy variable with the `format_dates` NatureCounts function. 
in.data<-format_dates(in.data)

```

We can then filter the data based on the Analysis Parameters file. This step will filter out the surveys which are done outside the approparite weather conditions or outside the survey window. We will create a before and after summary for comparison, so that it is clear how much data is lost due to this cleaning process. 

Gather the parameters you need from the analysis parameters file

```{r dataclean, echo = TRUE, message = FALSE}

#loop through each row in the analysis parameter files is doing multiple protocol
for(k in 1:nrow(anal.param)) {

#k<-11# manualy specify the row in the Analysis Parameters file to be run. 

max.yr <- 2019 #already specified in the setup, but doing it again here
protocol_id<-anal.param[k,"protocol_id"]
collection<-anal.param[k, "collection"]
min.yr <- anal.param[k,"min.year"] 
min.doy <- anal.param[k,"min.doy"]
max.doy <- anal.param[k,"max.doy"]
temp <- anal.param[k,"temp"]

#subset data based on protocol ID
dat<-in.data %>% filter(ProtocolCode == protocol_id)
dat$ObservationCount[is.na(dat$ObservationCount)] <- 0

#reassign ON route to QC
if(protocol_id=="35"){
dat$StateProvince[dat$StateProvince  == "ON"]  <- "QC"  
}

#Subset data to specified year range
#max.yr <- anal.param[k,"max.year"] #manually set to 2019 at this time
dat <- dat %>% filter(survey_year >= min.yr & survey_year <= max.yr)

#Create data summary prior to filtering
#nroute/year
route1<-dat %>% group_by(survey_year) %>% summarize(nroute_unfiltered=n_distinct(RouteIdentifier))

#nstop/route
stops<-dat %>% group_by(RouteIdentifier, survey_year) %>% summarize(nstops=n_distinct(SiteCode))
stops<-cast(stops, RouteIdentifier~survey_year, value="nstops")
write.csv(stops, paste(out.dir, collection, ".", protocol_id, ".StopsSummary.csv", sep=""))

#Speices/year level summary
stat.sp1<-dat %>% group_by(survey_year, CommonName) %>% summarize(count=sum(ObservationCount))

if(protocol_id=="44d"){
  stat.sp1 <- stat.sp1 %>% filter(survey_year!=2002) %>% filter(survey_year!=2010) %>% droplevels()
} #end if protocol id 44d

if(protocol_id=="100"){
 stat.sp1 <- stat.sp1 %>% filter(survey_year!=2000) %>% droplevels()
}

sp.dat1<-cast(stat.sp1, CommonName~survey_year, value="count")
sp.dat1$Filter<-"unfilter"

#Filtering the data to standardized detectability, and remove routes and speices that don't mean minimum cutoff requirements. In this step we also create the zero fill matrix. 

#Subset data to specified doy range
#This really butched some of the datasets. I think we can justify turning this filter off. 
#dat <- dat %>% filter(doy >= min.doy & doy <= max.doy)

#Subset data to minimum temperature using the start temp at the beginning of the survey. We will also keep NAs and assume these dataset are ok. This will help us preserve data. Could change moving forward. 
dat <- dat %>% filter(EffortMeasurement3 >= temp | is.na(EffortMeasurement3))

#Subset data to remove data collected when start precip > 1, start wind > 3 or noise > 3. We will also keep NAs and assume they are collected under appropriate conditions to help preserve data. Could change moving forward. Currently removed noise.   
dat<-dat %>% filter(EffortMeasurement1 <= 3 | is.na(EffortMeasurement1)) #Start wind
dat<-dat %>% filter(EffortMeasurement5 <= 1 | is.na(EffortMeasurement5)) #Start precip
#dat<-dat %>% filter(EffortMeasurement11 <= 3 | #is.na(EffortMeasurement11)) #Noise Level

#Now we want to remove repeat counts of individuals between stops. Not all region record this, but many do.
dat<-dat %>% filter(is.na(EffortMeasurement14)|EffortMeasurement14==0) #Repeats removed

#Now we impose some minimum data cut offs 
#Route must be run a minmum of 5 years
route.year<-dat %>% group_by(RouteIdentifier) %>% summarize(n_years_route=n_distinct(survey_year))
dat<-left_join(dat, route.year, by="RouteIdentifier")
dat<-dat %>% filter(n_years_route>=5) %>% 
  select(-n_years_route) 

#The number of stops per route >= 5
stop.year<-dat %>% group_by(RouteIdentifier, survey_year) %>% summarize(n_stop=n_distinct(SiteCode))
dat<-left_join(dat, stop.year, by=c("RouteIdentifier", "survey_year"))
dat<-dat %>% filter(n_stop>=5) 

#Because in the early years the Ontario owls program ran upwards of 4 surveys/ route/ year, the douplicates will need to be removed
if(collection=="ONOWLS"){
  
#since there is no window field we need to select the last survey date for those that have multiple surveys per rount within a year

#first filter by the min and max survey doy to capture the surveys that are done consistently with current protocol
test <- dat %>% filter(doy >= min.doy & doy <= max.doy)
  
#group by year and route to see unique survey months/days
test <- test %>% group_by(SurveyAreaIdentifier, survey_year)%>%
  summarize(sample_events=n_distinct(survey_month))

#merge back with main dataframe 
dat<-left_join(dat, test, by= c("SurveyAreaIdentifier", "survey_year"))

#subset out the data with sample_events>1
subowls<-subset(dat, sample_events>1)

#want to select the min day for each Year SurveryAreaIdentifier combo with more then one sampling event. 
minday<-subowls %>% group_by(SurveyAreaIdentifier, survey_year)%>%
  summarize(minday=min(doy))

#merge back with main data, elimiante the unwanted data, select needed columns
dat<-left_join(dat, minday, by= c("SurveyAreaIdentifier", "survey_year"))
dat$check<-dat$minday-dat$doy
dat$check[is.na(dat$check)] <- 0
dat<-dat%>%filter(check==0)
dat<-dat%>% select(-sample_events, -minday, -check)

} # end if ONOWLS 

#Now that the surveys are removed that were done in inapproparite envrionmnetal conditions and routes are removed that were not run in enough years we can make a list of unique sampling events to use for zero-filling species-specific dataframes. We don't zero-fill the entire dataframe for space issues. Note: Since our response variable in the analysis is counts at the Route level, the sampling event is a route within a year. If the analysis is changes to be at the stop level, this will need to be changed.   

event.data <-NULL #clear old
event.data <- dat %>%
  select(RouteIdentifier, survey_year, CollectorNumber, n_stop) %>%   
  distinct() %>%
  ungroup() %>%
  as.data.frame()

#Create data summary after creating the sampling events data layer.
#nroute/year
route2<-dat %>% group_by(survey_year) %>% summarize(nroute_filter=n_distinct(RouteIdentifier))

#nstop/route
stops2<-dat %>% group_by(RouteIdentifier, survey_year) %>% summarize(nstops=n_distinct(SiteCode))
stops2<-cast(stops2, RouteIdentifier~survey_year, value="nstops")

#Species must be detected on at least half of all years surveyed.  
min.yrs.detect<-trunc(length(unique(dat$survey_year))/2)

df.nyears <- NULL #clear old
df.nyears <- dat %>%
  filter(ObservationCount > 0) %>%
  select(survey_year, CommonName) %>%
  distinct() %>%
  group_by(CommonName) %>%
  summarize(nyears = n()) %>%
  filter(nyears >= min.yrs.detect)%>%
  as.data.frame()

dat <- left_join(df.nyears, dat, by = c("CommonName")) %>%  select(-nyears)

#Species must be detected with a mean of ~5 per years to be included in the analysis (this is actually set to 4 because some prtocol are very close at 4.8) 

df.abund <- NULL #clear old
df.abund <- dat %>%
  group_by(CommonName, survey_year) %>%
  summarize(count = sum(ObservationCount)) %>% 
  summarize(meanCount = mean(count, na.rm = TRUE)) %>% 
  filter(meanCount >= 4)%>%
  as.data.frame()

dat <- left_join(df.abund, dat, by = c("CommonName")) %>%
  select(-meanCount)%>%
  as.data.frame()

#Species/year level summary after filters 
stat.sp2<-dat %>% group_by(survey_year, CommonName) %>% summarize(count=sum(ObservationCount))
sp.dat2<-cast(stat.sp2, CommonName~survey_year, value="count")
sp.dat2$Filter<-"filter"

#now we want to just keep the observation records that were made in the first 2 min silent listening period (ObservationCount2 or ObservationCount3)

dat$ObservationCount2<-ifelse(is.na(dat$ObservationCount2), 0, 1)
dat$ObservationCount3<-ifelse(is.na(dat$ObservationCount3), 0, 1)
dat<- dat %>% mutate(ObservationCount1 = ifelse (ObservationCount2==1 | ObservationCount3==1, 1, 0))

#Now that we have the 'new' ObservationCount1 variable, we want to ensure that the minimum abundance filters are reapplied

#Species must be detected on at least half of all years surveyed.  
df2.nyears <- NULL #clear old
df2.nyears <- dat %>%
  filter(ObservationCount1 > 0) %>%
  select(survey_year, CommonName) %>%
  distinct() %>%
  group_by(CommonName) %>%
  summarize(nyears = n()) %>%
  filter(nyears >= min.yrs.detect)%>%
  as.data.frame()

dat1 <- left_join(df2.nyears, dat, by = c("CommonName")) %>%  select(-nyears)

#Species must be detected with a mean of ~5 per years to be included in the analysis (this is actually set to 4 because some prtocol are very close at 4.8) 

df2.abund <- NULL #clear old
df2.abund <- dat1 %>%
  group_by(CommonName, survey_year) %>%
  summarize(count = sum(ObservationCount1)) %>% 
  summarize(meanCount = mean(count, na.rm = TRUE)) %>% 
  filter(meanCount >= 4)%>%
  as.data.frame()

dat1 <- left_join(df2.abund, dat1, by = c("CommonName")) %>%
  select(-meanCount)%>%
  as.data.frame()

#Species/year level summary after Observation1 created  
stat.sp3<-dat %>% group_by(survey_year, CommonName) %>% summarize(count=sum(ObservationCount1))
sp.dat3<-cast(stat.sp3, CommonName~survey_year, value="count")
sp.dat3$Filter<-"2 min obs"

#Bind and write tables to output file
#the following are causing issue with the cbind.
#some issues with the BC FLAM data. Not enought records in 2010 to meet min data requirements

if(protocol_id=="44d"){
  route1 <- route1 %>% filter(survey_year!=2002) %>% filter(survey_year!=2010) %>% droplevels()
  } #end if protocol id 44d

if(protocol_id=="100"){
 route1 <- route1 %>% filter(survey_year!=2000) %>%  droplevels()
} #end if protocol id 100

route<-full_join(route1, route2, by="survey_year")
write.csv(route, paste(out.dir, collection, ".", protocol_id, ".RouteYearSummary.csv", sep=""))

sp.dat<-rbind(sp.dat1, sp.dat2, sp.dat3)
sp.dat<-arrange(sp.dat, CommonName)
write.csv(sp.dat, paste(out.dir, collection, ".", protocol_id, ".SpYearSummary.csv", sep=""))

#print the final data to file
#write.csv(dat, paste(out.dir, collection, ".", protocol_id, ".ObservationCount.csv", sep=""))
#write.csv(dat1, paste(out.dir, collection, ".", protocol_id, ".ObservationCount1.csv", sep=""))

#print the event.data to file 
#write.csv(event.data, paste(out.dir, collection, ".", protocol_id, ".EventData.csv", sep=""))

} #end loop

```





