---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Getting Started {#DataManip3}

```{r tidyr3, echo = FALSE, message = FALSE, warning = FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy = FALSE)

```

## Setup

The setup files will get you ready for your analysis. It loads you libraries, needed tables, created directories for your outputs, and assigns some parameter values. 

Also assign max year of the analysis. 

```{r setup}

min.yr <-2018
max.yr <- 2023 #needs changes with each analysis
ID<- "dethier" #change this to your user ID. You will be prompted for a password. 

source("00-setup.R")

```

## Choose a collection to be downloaded and prepare the provincial basemap

```{r downloaddata, echo = TRUE, message = FALSE}

#list the collections
list.coll<-unique(anal.param$collection)
list.prot<-unique(anal.param$protocol_id)

#select your collection manually for data cleaning
collection <- "MBOWLS"

```

## Output Tables

Create the output tables. This only needs to be done once per provincal analysis

```{r data manip}

source("i-output.R") #Total count period output tables
source("i-output2.R") #Silent listening period output tables

```


## Data manipulations 

```{r data manip}

source("ii-datamanip.R") #Total count period
source("ii-datamanip2.R") #Silent listening period

```

## Data Analysis iCAR

```{r analysis}

source("iii-icar.R")  #Total count period, including a covariate for number of stops per route and random effect for protocol_id to control for unexplained variation due to differences in protocol. 

```

## Data Analysis SPDE

```{r analysis}


#pull up the provincial basemap manually
qq <- rnaturalearth::ne_states(country = "canada", returnclass = "sf") %>% st_transform(epsg6703km) 
#qq <- filter(canada, name %in% c("British Columbia"))

source("iv-spde.R")

```

## Plot Trends 

```{r Plot trends}

source("v_plots.R")

```

## Plot Sum Stats and Visuals

```{r final outputs}

library(sf)
library(mapview)
library(ggmap) 
library(mapproj)
register_stadiamaps("1cdeb586-3f5a-4ef9-813b-02375efa9f21") 

##Plot all locations
loc <- read.csv("output/AllEventsNOS.csv")
loc<-loc %>% select(RouteIdentifier, latitude, longitude, survey_year) %>% distinct()

min.yr<-2008
max.yr<-2021

loc<-loc %>% filter(survey_year>=min.yr & survey_year<=max.yr)
test<-loc %>% select(RouteIdentifier) %>% distinct()

#View Data
sf<-st_as_sf(loc, coords = c("longitude", "latitude"), crs = 4326)
mapview(sf, map.types = "Esri.WorldImagery")

#Routes with more than 10 years of data
loc2<-loc %>% group_by(RouteIdentifier, latitude, longitude) %>% summarise(n=n_distinct(survey_year))
loc2<-loc2 %>% filter(n>=10)

#View Data
sf2<-st_as_sf(loc2, coords = c("longitude", "latitude"), crs = 4326)
mapview(sf2, map.types = "Esri.WorldImagery")


#compile the province year summaries

##Combine data outputs into a single table for data processing

prov_sum <- list.files(path = "C:/Users/dethier/Documents/ethier-scripts/National-NOS/output/",  # Identify all CSV files
                     pattern = "*ProvinceSummary.csv", full.names = TRUE) %>% 
  lapply(read_csv) %>%           # Store all files in list
  bind_rows                      # Combine data sets into one data set 

n<-nrow(prov_sum)
prov_sum<-prov_sum[2:n,]
prov_sum<-na.omit(prov_sum)
prov_sum$value<-1
prov_sum<-cast(prov_sum, StateProvince ~ Species)
write.csv(prov_sum, "output/AllProvinceSummaryNOS.csv")

```