---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Getting Started {#DataManip3}

```{r tidyr3, echo = FALSE, message = FALSE, warning = FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy = FALSE)

```

## Setup

The setup files will get you ready for your analysis. It loads you libraries, needed tables, created directories for your outputs, and assigns some parameter values. 

Also assign max year of the analysis. 

```{r setup}

max.yr <- 2023 #needs changes with each analysis
ID<- "dethier" #change this to your user ID. You will be prompted for a password. 

```


#Output Tables

Create the output tables. This only needs to be done once per analysis

```{r data manip}

source("i-outputs.R")

```

#Data Download/ Select Collection

The NOS data are downloaded using the `naturecounts` R package. Detailed instruction on how to use the `naturecounts` R pacakge can be found in [NatureCounts: An Introductory Tutorial](link%20to%20be%20provided) or on the [github](https://birdstudiescanada.github.io/naturecounts/index.html) repository. These instructions are not repeated here.

To access the NOS collections you must [sign up](https://www.birdscanada.org/birdmon/default/register.jsp) for a **free** account and request permission from the data custodian to access each `collection`. Once assess has been granted to can retrieve the data using the `request_id` provided. Also, don't forget to enter your personal `username` below. You will be promoted for your password.

There are many build in options which enable you to select temporal and/or spatial filters prior to download. If you want targeted data, you should add filters to the code chunk below. Because we will need additional auxiliary data from the [Bird Monitoring Data Exchange] (<https://www.birdscanada.org/birdmon/default/nc_bmde.jsp>) (BMDE) table in NatureCounts, we will want to change the `field_set` to "extend" and the `username` to your own account.

Choose a collection to be downloaded. 
```{r downloaddata, echo = TRUE, message = FALSE}

list.coll<-unique(anal.param$collection)

#select you collection manually
collection <- "ONOWLS"

```

## Data manipulations 

A file was created that outlines analysis parameters for various
Provinces.

**Alberta still needs assigned a protocol_id. Need to double check which
provinces have 2 listening periods in the database**

Column definitions

-   collection: each province/region has a unique collection ID

-   protocol_id: protocol vary between provinces, and some
    provinces/regions have several protocols within them. However, not
    all protocol_id are specified in this field and need to be parsed
    out using auxiliary data (e.g., BCOWLS uses the BCregions to
    differentiate protocol_id)

-   min.year: start year of survey

-   max.year: end year of survey, or set to NA if ongoing

-   temp: min temperature cut off for survey

-   min.doy: the earliest day of the year a survey should be run

-   max.doy: the latest day of the year a survey should be run

-   interval: indicates if the two minute silent listening periods is
    recorded in 1 or 2 data columns. If 2 data columns then these
    datasets are ideal for modelling the detection process

-   obs: the observation count type that should be used for the national
    analysis.

#At the moment, the data is manipualted provinically.   
  
```{r data manip}

source("ii_datamanip.R")

```

## Data Analysis iCAR

```{r analysis}

source("iii_analysis.R")

```

## Data Analysis SPDE

```{r analysis}

source("iii_analysis.R")

```



## Plot Trends 

```{r Plot trends}

source("iii_plottrend.R")

```

