---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Mapping {#Map4}

```{r tidyr4, echo = FALSE, message = FALSE, warning = FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy = FALSE)

```

Here we are going to map the Sites on a Route using the latitude and longitude for each province as part of the QAQC for the owls data. 

Load the packages you will need: 
```{r loadpackages2, echo = FALSE, message = FALSE}

library(rgdal)
library(sf)
library(ggmap)
library(leaflet)
library(mapview)
library(rnaturalearth)

```

Code to create Provincial plots through time using the filtered data

```{r plotprov, echo = TRUE, message = FALSE}

#Will want to loop through `collection` when the data are ready
prov<-unique(in.data$collection)

for(t in 1:length(prov)) {

 t<-13 #for testing individual provinces
  
plot.data<-subset(in.data, collection==prov[t])

plot.data<-plot.data %>% select(RouteIdentifier, survey_year, latitude, longitude) %>% distinct()

plot.data$survey_year<-as.factor(plot.data$survey_year)
plot.data<-plot.data %>% filter (!is.na(latitude), !is.na(longitude))

#Create a spatial dataframe

plot_sf <- st_as_sf(plot.data, coords = c("longitude", "latitude"), crs = 4326)

#Create the base layer map
map <- get_stamenmap(bbox = as.numeric(st_bbox(plot_sf)), zoom = 2)

#Create a new plot for each year
plot<-ggmap(map) + 
  #geom_point(data = plot.data, aes(x = longitude, y = latitude, colour = survey_year))+
  geom_point(data = plot.data, aes(x = longitude, y = latitude))+
  facet_wrap( ~ survey_year, ncol=5)

  
print(plot)

}
```

Code to create National plots using Events data

```{r plotnational, echo = TRUE, message = FALSE}

Events<-read.csv("output/Events.csv")

#use full dataset (could also include site code if you want to have the stop on the route shown. I don' think this is necessary at the national scale. Just one point per route)
plot.data<-Events %>% select(RouteIdentifier, survey_year, latitude, longitude) %>% distinct()

#remove NA in lat and long
plot.data<-subset(plot.data, !is.na(latitude)) 
plot.data<-subset(plot.data, !is.na(longitude)) 
plot.data$survey_year<-as.factor(plot.data$survey_year)

#remove outliers. Should be fixed in underlying data. 
plot.data <- plot.data %>%filter(!(RouteIdentifier %in% c("ON0233", "ON0328", "PE010", "QC062", "ON0099"))) %>% filter(latitude != "MB")

#Create a spatial dataframe
plot_sf <- st_as_sf(plot.data, coords = c("longitude", "latitude"), crs = 4326)

#Load map of Canada
canada <- ne_states(country = "canada", returnclass = "sf") %>%
  st_transform(3347)

#plot data to the map
plot<-ggplot() +
  theme_bw() +
  geom_sf(data = canada) +
  geom_sf(data = plot_sf, size = 0.5)

print(plot)

#Interactive Map
#range of year = 1991, 2020
#mapview(plot_sf, zcol = "survey_year", map.types = "Esri.WorldImagery")
```

Using the filtered data, we can also create maps of species detection across the nation. 

```{r sp maps}

Owls<-read.csv("output/Owls.csv")

#Filter out non-target species. 
obs <- Owls %>% filter (CommonName %in% c("Barred Owl", "Boreal Owl", "Great Gray Owl", "Great Horned Owl", "Northern Saw-whet Owl", "Eastern Screeh-Owl", "Western Screech-Owl", "Long-eared Owl", "Short-eared Owl", "Northern Hawk Owl", "Northern Pygmy-Owl", "Snowy Owl", "Barn Owl", "Flammulated Owl"))

obs<-obs %>% filter(RouteIdentifier!="NS050") %>% droplevels()

#species by protocol summary. Check for misclassifications
sp.prot<-obs %>% select(CommonName, collection, ProtocolCode, StateProvince) %>% distinct()
sp.prot$count<-"x"
sp.prot<-cast(sp.prot, collection + StateProvince+ProtocolCode~CommonName, value="count")

#write the table that detail the speices that will be included in the national assessment. 
write.csv(sp.prot, "output/Species_by_Protcol.csv")

#plot by species
plot.data<-dat %>% select(RouteIdentifier, survey_year, latitude, longitude, CommonName) %>% distinct()

#remove NA in lat and long
plot.data<-subset(plot.data, !is.na(latitude)) 
plot.data<-subset(plot.data, !is.na(longitude)) 
plot.data$survey_year<-as.factor(plot.data$survey_year)

#Create a spatial dataframe
plot_sf <- st_as_sf(plot.data, coords = c("longitude", "latitude"), crs = 4326)

#Create the base layer map
map <- get_stamenmap(bbox = as.numeric(st_bbox(plot_sf)), zoom = 2)

#Create a new plot for each year
plot<-ggmap(map) + 
  #geom_point(data = plot.data, aes(x = longitude, y = latitude, colour = survey_year))+
  geom_point(data = plot.data, aes(x = longitude, y = latitude))+
  facet_wrap( ~ CommonName, ncol=5)


print(plot)

```