---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Mapping {#Map1}

```{r tidyr4, echo = FALSE, message = FALSE, warning = FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy = FALSE)

```

Here we are going to map the Sites on a Route using the latitude and longitude for each province as part of the QAQC for the owls data. 

Load the packages you will need: 
```{r loadpackages2, echo = FALSE, message = FALSE}

library(rgdal)
library(sf)
library(ggmap)
library(leaflet)
library(mapview)
library(rnaturalearth)

```

Code to create Provincial plots through time using the unfiltered data

```{r plotprov, echo = TRUE, message = FALSE}

#Will want to loop through `collection` when the data are ready
prov<-unique(in.data$collection)

for(t in 1:length(prov)) {

#  t<-13 #for testing individual provinces
  
plot.data<-subset(in.data, collection==prov[t])

plot.data<-plot.data %>% select(RouteIdentifier, survey_year, latitude, longitude) %>% distinct()

plot.data$survey_year<-as.factor(plot.data$survey_year)
plot.data<-plot.data %>% filter (!is.na(latitude), !is.na(longitude))

#Create a spatial dataframe

plot_sf <- st_as_sf(plot.data, coords = c("longitude", "latitude"), crs = 4326)

#Create the base layer map
map <- get_stamenmap(bbox = as.numeric(st_bbox(plot_sf)), zoom = 5)

#Create a new plot for each year
plot<-ggmap(map) + 
  geom_point(data = plot.data, aes(x = longitude, y = latitude, colour = survey_year))+
  facet_wrap( ~ survey_year, ncol=5)

  
print(plot)

}
```

Code to create National plots 

```{r plotnational, echo = TRUE, message = FALSE}

#use full dataset (could also include site code if you want to have the stop on the route shown. I don' think this is necessary at the national scale. Just one point per route)
plot.data<-in.data %>% select(ObservationCount, RouteIdentifier, survey_year, latitude, longitude) %>% distinct()

#remove NA in lat and long
plot.data<-subset(plot.data, !is.na(latitude)) 
plot.data<-subset(plot.data, !is.na(longitude)) 
plot.data$survey_year<-as.factor(plot.data$survey_year)

#remove outliers. Should be fixed in underlying data. 
plot.data <- plot.data %>%filter(!(RouteIdentifier %in% c("ON0233", "ON0328", "PE010", "QC062", "ON0099")))

#Create a spatial dataframe
plot_sf <- st_as_sf(plot.data, coords = c("longitude", "latitude"), crs = 4326)

#Load map of Canada
canada <- ne_states(country = "canada", returnclass = "sf") %>%
  st_transform(3347)

#plot data to the map
plot<-ggplot() +
  theme_bw() +
  geom_sf(data = canada) +
  geom_sf(data = plot_sf, size = 0.5)

print(plot)

#Interactive Map
#range of year = 1991, 2020
#mapview(plot_sf, zcol = "survey_year", map.types = "Esri.WorldImagery")
```



## Set plot and map themes {#Setup2.5}

This may be moved to a later part of the script. Not need here.

```{r plottheme echo=FALSE, message=FALSE}
# plot theme
theme_timeseries <- function (base_size = 11, base_family = "") {
  theme_grey(base_size = base_size, base_family = base_family) %+replace%
    theme(panel.background = element_rect(fill = "white", colour = NA),
          panel.border = element_rect(fill = NA, colour = "grey20"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(size = rel(0.9), angle = 0),
          axis.text.y = element_text(size = rel(0.9), angle = 0),
          strip.background = element_rect(fill = "grey80"),
          legend.key = element_rect(fill = "white", colour = NA),
          plot.title = element_text(size=14, hjust = 0.5,
                                    margin=margin(t=5, b=10)),
          legend.position="right",
          complete = TRUE)
}; theme_set(theme_timeseries())

# map theme
theme_map <- function(base_size = 9, base_family = "") {
  theme_bw(base_size = base_size, base_family = base_family) %+replace%
    theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid = element_blank(),
          panel.spacing = unit(0, "lines"),
          plot.background = element_blank(),
          legend.background=element_rect(fill=NA, colour=NA),
          legend.direction="vertical",
          legend.key=element_rect(fill=NA, colour="white"),
          legend.text.align=1,
          legend.text = element_text(size=9),
          legend.title=element_text(hjust=0, size=11),
          legend.justification=c(0, 0.5),
          plot.title = element_text(size=14, hjust = 0.7))
}

options(scipen=9999999)
options(max.print=99999)
options(stringsAsFactors=F)

```
